---
title: "tidy-model"
author: "Aleph Girmaye"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) ; library(modeldata)
```

## Introduction

This R Markdown document demonstrates my abilities to use models for data analysis using a data set collected on crickets.

```{r load data, include=FALSE}
data(crickets, package = "modeldata")
names(crickets)

```

## What is that sound?
In this report, we examine what factors predict a cricket's chirp rate.

```{r summary , echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) +
   geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of chirp rate (per min.)")
```

## Temperature affects chirp rate.

```{r temp, echo = FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)
summary.lm(temp_lm)

```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species , echo=FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

  species_lm <- lm(rate ~ temp + species, crickets)
  summary.lm(species_lm)

```

Compared to the first visual, this scatter plot and summary model shows a better correlation between chirp rates and temperature by splitting the data points by species. This shows us that "exclamationis" has a consistently higher chirp rate than "niveus" at the same temperature. Also based  this model we get a better account of the data with a 99 % of the variance.

## Interactions

```{r species historgram, echo=FALSE}
ggplot(crickets, aes(x = rate, fill = species)) +
  geom_histogram(position = 'identity', alpha = 0.7, bins = 8) +
  ggtitle("Distribution of the chirp rate of crickets.") +
  xlab('Chirp rate (per min.)')

```


```{r intersection , echo=FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

  species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets) 
  summary.lm(species_x_temp_lm)

  anova(species_lm, species_x_temp_lm)

```
After checking the intersection between temperature and species by adding a temperature-species interaction we see that the relationship between temperature and chirp rate does not differ meaningfully between species. Because of this it s better to stick with the  species model.

```{r load_diamonds, include = FALSE}
data(diamonds, package = "ggplot2")
names(diamonds)

```


```{r }

```
